---
title: "Phase 1 Stage 3 community-composition analyses"
subtitle: "A tutorial for db-RDA, Hellinger RDA, and principal response curves"
author: "Trophosome project"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
params:
  data_dir: "experiments/work/trophosome/p01-neutral-feedback/analysis/s03-parameter-map-community-wave3-g100-derived"
  nperm: 9999
  random_seed: 20260908
  run_models: true
---

# Purpose

This notebook analyses **whole environmental-community composition**. It uses
three complementary methods:

1. **distance-based redundancy analysis (db-RDA)** asks whether experimental
   factors explain total-variation differences among passage-100 communities;
2. **Hellinger RDA** asks the same endpoint questions in a transformed lineage-
   abundance table and makes lineage contributions easier to inspect;
3. **principal response curves (PRC)** ask how selected treatment trajectories
   depart from a declared reference trajectory through passages 0--100.

The response columns are the 100 frozen ancestral lineages. Every mutant is
traced back to its ancestral root, and its abundance is added to that root.
This makes columns comparable across simulations. These analyses therefore
describe **ancestral-lineage composition**, not mutant richness or the survival
of individual mutant strains.

The endpoint methods and PRC answer different questions. The endpoint methods
compare communities at passage 100. PRC uses the complete trajectory and asks
when and how treatments diverge from a reference. PRC should be run on a small,
biologically coherent comparison, not on all 81 Stage 3 cells at once.

# 1. Set up R

Install the packages once if necessary. This chunk is deliberately not run
when the document is knitted.

```{r install-packages, eval=FALSE}
install.packages(c("vegan", "permute", "ggplot2", "knitr", "rmarkdown"))
```

```{r setup, include=FALSE}
required_packages <- c("vegan", "permute", "ggplot2", "knitr")
missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]
if (length(missing_packages) > 0) {
  stop(
    "Install these R packages before knitting: ",
    paste(missing_packages, collapse = ", ")
  )
}

knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 5,
  dpi = 120
)

# Find the repository whether knitting starts from the repository root or docs/.
find_repository <- function(start = getwd()) {
  candidate <- normalizePath(start, mustWork = TRUE)
  repeat {
    if (
      file.exists(file.path(candidate, "README.md")) &&
      dir.exists(file.path(candidate, "experiments", "work", "trophosome"))
    ) {
      return(candidate)
    }
    parent <- dirname(candidate)
    if (identical(parent, candidate)) {
      stop("Could not locate the Modeling_trophosome repository.")
    }
    candidate <- parent
  }
}

repository <- find_repository()
derived <- file.path(repository, params$data_dir)
analysis_output <- file.path(derived, "r-community-analysis")
dir.create(analysis_output, recursive = TRUE, showWarnings = FALSE)
set.seed(params$random_seed)
```

The permutation count defaults to `r params$nperm`. During code development,
set `nperm` to 99 for speed. Restore 9,999 or more for reported analyses and
record the random seed.

# 2. Read the compiled tables

The compiler creates one master endpoint triplet and one additional trajectory
table:

- **X** contains experimental variables and provenance;
- **Y** contains passage-100 ancestral-lineage frequencies;
- **Y-prime** contains pairwise total-variation distances calculated from Y;
- the **PRC trajectory table** contains the same lineage frequencies at every
  passage from 0 through 100.

```{r read-endpoint-tables}
paths <- c(
  X = file.path(derived, "x-explanatory-g100.tsv"),
  Y = file.path(derived, "y-ancestral-frequencies-g100.tsv"),
  D_TV = file.path(derived, "yprime-tv-g100.tsv"),
  PRC = file.path(derived, "prc-ancestral-trajectories-g0-g100.tsv.gz")
)

missing_files <- paths[!file.exists(paths)]
if (length(missing_files) > 0) {
  stop(
    "The compiler output is incomplete. Missing: ",
    paste(basename(missing_files), collapse = ", ")
  )
}

X <- read.delim(paths[["X"]], check.names = FALSE, stringsAsFactors = FALSE)
Y_table <- read.delim(
  paths[["Y"]], check.names = FALSE, stringsAsFactors = FALSE
)
D_table <- read.delim(
  paths[["D_TV"]], check.names = FALSE, stringsAsFactors = FALSE
)

Y <- as.matrix(Y_table[, -1, drop = FALSE])
storage.mode(Y) <- "double"
rownames(Y) <- Y_table$sample_id

D_TV_matrix <- as.matrix(D_table[, -1, drop = FALSE])
storage.mode(D_TV_matrix) <- "double"
rownames(D_TV_matrix) <- D_table$sample_id
colnames(D_TV_matrix) <- names(D_table)[-1]
```

The compressed PRC table is larger, so it is read in a cached chunk. A second
knit can then reuse the cached result.

```{r read-prc-table, cache=TRUE}
PRC_table <- read.delim(
  gzfile(paths[["PRC"]]),
  check.names = FALSE,
  stringsAsFactors = FALSE
)
lineage_columns <- grep("^ancestral_[0-9]{3}$", names(PRC_table), value = TRUE)
```

# 3. Audit before analysing

These checks intentionally stop the notebook if row order, dimensions, or
frequency sums are wrong. Silent misalignment would produce a convincing but
incorrect ordination.

```{r audit-inputs}
stopifnot(!anyDuplicated(X$sample_id))
stopifnot(identical(X$sample_id, Y_table$sample_id))
stopifnot(identical(X$sample_id, D_table$sample_id))
stopifnot(identical(X$sample_id, names(D_table)[-1]))
stopifnot(identical(rownames(D_TV_matrix), colnames(D_TV_matrix)))
stopifnot(ncol(Y) == 100L, length(lineage_columns) == 100L)
stopifnot(all(Y >= 0), all(abs(rowSums(Y) - 1) < 1e-12))
stopifnot(all(D_TV_matrix >= 0), all(D_TV_matrix <= 1 + 1e-12))
stopifnot(max(abs(D_TV_matrix - t(D_TV_matrix))) < 1e-12)
stopifnot(max(abs(diag(D_TV_matrix))) < 1e-12)

# Recalculate a small sample of distances independently.
check_indices <- unique(round(seq(1, nrow(Y), length.out = min(12, nrow(Y)))))
for (i in check_indices) {
  for (j in check_indices) {
    expected <- 0.5 * sum(abs(Y[i, ] - Y[j, ]))
    stopifnot(abs(D_TV_matrix[i, j] - expected) < 1e-12)
  }
}

# Each analysis population must contribute exactly passages 0,...,100 to PRC.
stopifnot(!anyDuplicated(PRC_table$trajectory_sample_id))
stopifnot(all(PRC_table[, lineage_columns] >= 0))
stopifnot(
  all(abs(rowSums(PRC_table[, lineage_columns, drop = FALSE]) - 1) < 1e-12)
)
trajectory_sizes <- table(PRC_table$population_sample_id)
stopifnot(all(trajectory_sizes == 101L))
trajectory_passages <- split(
  PRC_table$passage, PRC_table$population_sample_id, drop = TRUE
)
stopifnot(all(vapply(
  trajectory_passages,
  function(z) identical(sort(as.integer(z)), 0:100),
  logical(1)
)))

data.frame(
  table = c("X endpoint", "Y endpoint", "Y-prime", "PRC trajectory"),
  rows = c(nrow(X), nrow(Y), nrow(D_TV_matrix), nrow(PRC_table)),
  columns = c(ncol(X), ncol(Y), ncol(D_TV_matrix), ncol(PRC_table))
)
```

# 4. Preserve the experimental design when subsetting

The Wave 3 master tables contain four experimental subsets. Always subset X first,
then use its `sample_id` values to subset Y and both axes of Y-prime.

```{r endpoint-helper}
as_logical_flag <- function(x) {
  tolower(as.character(x)) == "true"
}

subset_endpoint <- function(analysis_set, primary_only = TRUE) {
  keep <- X$analysis_set == analysis_set
  if (primary_only) {
    keep <- keep & as_logical_flag(X$include_primary_dbrda)
  }
  design <- X[keep, , drop = FALSE]
  ids <- design$sample_id
  if (length(ids) == 0) stop("The requested endpoint subset is empty.")
  if (!all(ids %in% rownames(Y))) stop("Some X sample IDs are absent from Y.")
  if (!all(ids %in% rownames(D_TV_matrix))) {
    stop("Some X sample IDs are absent from Y-prime.")
  }
  list(
    X = design,
    Y = Y[ids, , drop = FALSE],
    D_TV = stats::as.dist(D_TV_matrix[ids, ids, drop = FALSE])
  )
}

wave1 <- subset_endpoint("wave1_h_alpha_u")
wave2a <- subset_endpoint("wave2a_h_by_b")
wave2b <- subset_endpoint("wave2b_alpha_by_m")
wave3 <- subset_endpoint("wave3_bridge")

stopifnot(nrow(wave1$X) == 288L)
stopifnot(nrow(wave2a$X) == 144L)
stopifnot(nrow(wave2b$X) == 336L)
stopifnot(nrow(wave3$X) == 96L)
```

Wave 1's 12 no-return controls remain in the master tables but are not in its
primary (H \times \alpha \times u) factorial model. When \(\alpha=0\), host
abundance and mutation rate cannot act through host return, so attaching an
arbitrary H or u label to that control would distort the factorial test. Use
those controls for paired treatment-versus-control contrasts and selected PRCs.

## 4.1 Build the bridge-augmented subset

Wave 3 was designed to augment compatible neutral observations from Waves 1
and 2, rather than to be fitted as an isolated 16-cell factorial. Use only the
six seed blocks shared by Wave 3, keep mutation off, and represent each original
simulated population once. The last rule matters because some earlier
populations deliberately appear under more than one analysis-cell alias.

```{r bridge-augmented-helper}
bridge_seed_blocks <- sprintf("sb%04d", 1:6)
bridge_sets <- c(
  "wave1_h_alpha_u", "wave2a_h_by_b", "wave2b_alpha_by_m", "wave3_bridge"
)
bridge_keep <-
  X$analysis_set %in% bridge_sets &
  X$seed_block_id %in% bridge_seed_blocks &
  X$u == 0 &
  as_logical_flag(X$include_primary_dbrda)

bridge_design <- X[bridge_keep, , drop = FALSE]

# If one source appears through multiple analysis aliases, its biological
# predictors must agree before one alias is retained.
predictor_columns <- c("H", "B", "alpha", "m", "u")
aliases <- split(bridge_design, bridge_design$source_run_id, drop = TRUE)
aliases_agree <- vapply(
  aliases,
  function(z) nrow(unique(z[, predictor_columns, drop = FALSE])) == 1L,
  logical(1)
)
stopifnot(all(aliases_agree))

bridge_design <- bridge_design[
  !duplicated(bridge_design$source_run_id), , drop = FALSE
]
bridge_ids <- bridge_design$sample_id
bridge <- list(
  X = bridge_design,
  Y = Y[bridge_ids, , drop = FALSE],
  D_TV = stats::as.dist(
    D_TV_matrix[bridge_ids, bridge_ids, drop = FALSE]
  )
)

stopifnot(!anyDuplicated(bridge$X$source_run_id))
stopifnot(nrow(bridge$X) == 390L)
stopifnot(all(table(bridge$X$seed_block_id) == 65L))
table(bridge$X$analysis_set)
```

This is still one master X/Y/Y-prime release: `bridge` is only a checked view
of it in memory. No additional copy of the matrices is maintained.

# 5. Permutations: what is the independent replicate?

The independent replicate is a **simulated population identified by a seed
block**, not a lineage, host, or passage. The earlier panels use 12 matched seed
blocks; the bridge-augmented analysis uses the six blocks shared with Wave 3.
Endpoint permutations are therefore restricted within seed block. The seed
block is also conditioned out of the ordination so seed-specific starting/drift
differences are not attributed to the treatments.

```{r endpoint-model-helper}
prepare_design <- function(design) {
  design$seed_block_id <- factor(design$seed_block_id)
  design$mutation_enabled <- factor(
    as_logical_flag(design$mutation_enabled),
    levels = c(FALSE, TRUE),
    labels = c("mutation off", "mutation on")
  )
  design$log10_alpha <- ifelse(
    design$alpha_target > 0,
    log10(design$alpha_target),
    NA_real_
  )
  design$alpha_factor <- factor(
    design$alpha_target,
    levels = sort(unique(design$alpha_target))
  )
  design$m_factor <- factor(design$m, levels = sort(unique(design$m)))
  design$z_log10_H <- as.numeric(scale(log10(design$H)))
  design$z_log10_B <- as.numeric(scale(log10(design$B)))
  design$z_alpha <- as.numeric(scale(design$alpha))
  design$z_m <- as.numeric(scale(design$m))
  design
}

fit_endpoint_pair <- function(endpoint, right_hand_side) {
  design <- prepare_design(endpoint$X)
  D_TV <- endpoint$D_TV
  Y_hellinger <- vegan::decostand(endpoint$Y, method = "hellinger")

  # The formula environment must contain D_TV and Y_hellinger.
  db_formula <- stats::as.formula(
    paste("D_TV ~", right_hand_side, "+ Condition(seed_block_id)"),
    env = environment()
  )
  rda_formula <- stats::as.formula(
    paste("Y_hellinger ~", right_hand_side, "+ Condition(seed_block_id)"),
    env = environment()
  )

  list(
    design = design,
    permutations = permute::how(
      nperm = params$nperm,
      blocks = design$seed_block_id
    ),
    dbrda = vegan::dbrda(db_formula, data = design),
    hellinger_rda = vegan::rda(rda_formula, data = design)
  )
}

ordination_tests <- function(model, permutation_design) {
  set.seed(params$random_seed)
  global <- anova(model, permutations = permutation_design)
  set.seed(params$random_seed)
  sequential <- anova(
    model, by = "terms", permutations = permutation_design
  )
  set.seed(params$random_seed)
  marginal <- anova(
    model, by = "margin", permutations = permutation_design
  )
  list(
    global = global,
    sequential = sequential,
    marginal = marginal,
    adjusted_R2 = vegan::RsquareAdj(model)
  )
}
```

The global test asks whether the explanatory set matters at all. The sequential
(`by = "terms"`) table depends on the order of terms and should follow a
predeclared biological hierarchy. The marginal table asks what a term adds
after the other eligible terms; a main effect contained in an interaction is
not separately marginal in the usual sense. For formal claims about a specific
interaction, compare predeclared nested models rather than selecting terms
after seeing their P-values.

# 6. Endpoint db-RDA

db-RDA uses the compiled total-variation distance directly. Because each Y row
sums to one, this is numerically the same as Bray--Curtis dissimilarity here.
The principal output is the permutation test and explained variation; an
ordination plot is a visualization, not an independent hypothesis test.

The primary formulas are:

| Experiment | Formula | Reason |
|---|---|---|
| Wave 1 | `log10_H * log10_alpha * mutation_enabled` | tests abundance, feedback, mutation, and their interactions |
| Wave 2A | `log10_H * log10_B` | tests whether H and bottleneck B have independent or interactive effects |
| Wave 2B | `alpha_factor * m_factor` | preserves a potentially nonlinear surface and includes alpha = 0 |
| Wave 1--3 bridge | all four scaled main effects plus `H:alpha`, `H:B`, `alpha:m`, `alpha:B`, `H:m`, and `B:m` | uses the new bridge cells to estimate the predeclared pairwise interactions |

```{r fit-endpoint-models, eval=params$run_models}
fits <- list(
  wave1 = fit_endpoint_pair(
    wave1, "log10_H * log10_alpha * mutation_enabled"
  ),
  wave2a = fit_endpoint_pair(wave2a, "log10_H * log10_B"),
  wave2b = fit_endpoint_pair(wave2b, "alpha_factor * m_factor"),
  bridge = fit_endpoint_pair(
    bridge,
    paste(
      "z_log10_H + z_alpha + z_log10_B + z_m",
      "+ z_log10_H:z_alpha + z_log10_H:z_log10_B",
      "+ z_alpha:z_m + z_alpha:z_log10_B",
      "+ z_log10_H:z_m + z_log10_B:z_m"
    )
  )
)
```

Start by inspecting Wave 1. Repeat the same calls for `wave2a`, `wave2b`, and
the combined `bridge` analysis.

```{r test-wave1-dbrda, eval=params$run_models}
wave1_db_tests <- ordination_tests(
  fits$wave1$dbrda, fits$wave1$permutations
)
wave1_db_tests
```

```{r dbrda-site-plot, eval=params$run_models, fig.cap="Wave 1 passage-100 communities in the first two constrained db-RDA dimensions. Points are population replicates."}
site_plot <- function(model, design, colour_variable, title) {
  site_scores <- as.data.frame(
    vegan::scores(model, display = "sites", choices = 1:2, scaling = 2)
  )
  if (ncol(site_scores) < 2) {
    stop("The fitted model has fewer than two constrained axes.")
  }
  names(site_scores)[1:2] <- c("Axis1", "Axis2")
  site_scores$colour <- factor(design[[colour_variable]])
  ggplot2::ggplot(
    site_scores,
    ggplot2::aes(x = Axis1, y = Axis2, colour = colour)
  ) +
    ggplot2::geom_point(alpha = 0.75, size = 2) +
    ggplot2::theme_bw() +
    ggplot2::labs(title = title, colour = colour_variable)
}

site_plot(
  fits$wave1$dbrda,
  fits$wave1$design,
  "alpha_target",
  "Wave 1 db-RDA: colour shows host-return fraction"
)
```

Total variation is not guaranteed to be Euclidean, so a db-RDA can contain
negative unconstrained eigenvalues. Record their size. If they are appreciable,
repeat the fit with a declared correction such as `add = "lingoes"` and report
whether conclusions change; do not silently switch distances or corrections.

```{r inspect-negative-eigenvalues, eval=params$run_models}
all_eigenvalues <- vegan::eigenvals(fits$wave1$dbrda, model = "all")
negative_fraction <- sum(abs(all_eigenvalues[all_eigenvalues < 0])) /
  sum(abs(all_eigenvalues))
negative_fraction
```

# 7. Endpoint Hellinger RDA

The Hellinger transform is the square root of each lineage's relative
abundance. It reduces the leverage of very abundant lineages and makes
Euclidean RDA suitable for community-composition data containing many zeros.
It is a valuable companion analysis, not a replacement for the prespecified
total-variation db-RDA.

```{r test-wave1-hellinger, eval=params$run_models}
wave1_hellinger_tests <- ordination_tests(
  fits$wave1$hellinger_rda, fits$wave1$permutations
)
wave1_hellinger_tests
```

```{r hellinger-site-plot, eval=params$run_models, fig.cap="Wave 1 passage-100 communities in the first two constrained Hellinger-RDA dimensions."}
site_plot(
  fits$wave1$hellinger_rda,
  fits$wave1$design,
  "alpha_target",
  "Wave 1 Hellinger RDA: colour shows host-return fraction"
)
```

Lineage scores indicate which ancestral lineages contribute most strongly to
the displayed axes. Label only the largest scores; labelling all 100 lineages
usually makes an unreadable figure.

```{r top-lineage-scores, eval=params$run_models}
lineage_scores <- as.data.frame(
  vegan::scores(
    fits$wave1$hellinger_rda,
    display = "species",
    choices = 1:2,
    scaling = 2
  )
)
lineage_scores$lineage <- rownames(lineage_scores)
lineage_scores$magnitude <- sqrt(
  lineage_scores[[1]]^2 + lineage_scores[[2]]^2
)
head(lineage_scores[order(-lineage_scores$magnitude), ], 12)
```

# 8. Principal response curves

PRC is a special partial RDA of treatment-by-time effects after conditioning
on time. The first level of the treatment factor is the reference treatment,
and time must be an unordered factor. See the official
[vegan PRC documentation](https://vegandevs.github.io/vegan/reference/prc.html).

Here PRC uses Hellinger-transformed ancestral frequencies. This keeps its
community geometry consistent with the Hellinger RDA and limits domination by
the most abundant ancestral lineages. A raw-frequency PRC can be added as a
prespecified sensitivity analysis.

## 8.1 Select one coherent comparison

This example asks: **at migration m = 0.1 and H = 1,000, how do three positive
host-return levels depart from the matched no-return trajectory?** It selects
Wave 2B cells c0066, c0073, c0080, and c0087. Each treatment has one population
for every seed block and 101 repeated observations within that population.

```{r select-prc-comparison}
select_prc_comparison <- function(
  analysis_set, cell_labels, reference_cell
) {
  cells <- names(cell_labels)
  if (!reference_cell %in% cells) stop("Reference cell is not in cell_labels.")
  selected <- PRC_table[
    PRC_table$analysis_set == analysis_set & PRC_table$cell %in% cells,
    ,
    drop = FALSE
  ]
  if (nrow(selected) == 0) stop("The requested PRC subset is empty.")

  treatment_order <- c(reference_cell, setdiff(cells, reference_cell))
  selected$treatment <- factor(
    unname(cell_labels[selected$cell]),
    levels = unname(cell_labels[treatment_order])
  )
  selected$seed_block_id <- factor(selected$seed_block_id)
  selected$time <- factor(selected$passage, levels = 0:100, ordered = FALSE)
  selected <- selected[
    order(
      selected$seed_block_id,
      selected$population_sample_id,
      selected$passage
    ),
    ,
    drop = FALSE
  ]

  # Check the experimental unit: one complete population trajectory per
  # seed-block-by-treatment combination.
  population_index <- unique(selected[
    , c("population_sample_id", "seed_block_id", "treatment")
  ])
  stopifnot(all(table(
    population_index$seed_block_id, population_index$treatment
  ) == 1L))
  stopifnot(all(table(selected$population_sample_id) == 101L))
  stopifnot(levels(selected$treatment)[1] == cell_labels[[reference_cell]])
  selected
}

wave2b_m010_prc <- select_prc_comparison(
  analysis_set = "wave2b_alpha_by_m",
  cell_labels = c(
    c0066 = "no host return",
    c0073 = "alpha = 0.01",
    c0080 = "alpha = 0.1",
    c0087 = "alpha = 0.99"
  ),
  reference_cell = "c0066"
)

unique(wave2b_m010_prc[
  , c("cell", "treatment", "H", "alpha_target", "m", "u")
])
```

Other useful, separate PRCs include:

- Wave 1 host return at fixed H = 1,000 and u = 0: no-return c0021 versus
  c0035, c0037, c0039, and c0041;
- Wave 2A bottleneck size at fixed H = 1,000: c0055--c0058, with one declared
  B value as the reference;
- Wave 2B migration at fixed alpha = 0.1: c0077--c0083, with m = 0 as the
  reference;
- Wave 3 feedback × bottleneck at H = 1,000 and m = 0.1: c0091--c0096, using
  one declared alpha/B combination as the reference;
- Wave 3 host abundance × migration at alpha = 0.1 and B = 10: c0097--c0100;
  and
- Wave 3 bottleneck × migration at H = 1,000 and alpha = 0.1: c0101--c0104.

These answer different biological contrasts and should not be merged into a
single omnibus PRC.

## 8.2 Fit and test the PRC

Treatment labels were assigned to whole simulated populations. Therefore,
permutations must move **complete trajectories** between treatment labels
within each matched seed block. The 101 passages are repeated measurements,
not 101 independent replicates. `Within(type = "none")` preserves their order;
`Plots(...)` identifies the whole population trajectory; `blocks` prevents
exchange across seed blocks. The permutation grammar is documented by
[permute](https://gavinsimpson.github.io/permute/reference/how.html).

```{r fit-prc, eval=params$run_models}
prc_response <- as.matrix(
  wave2b_m010_prc[, lineage_columns, drop = FALSE]
)
storage.mode(prc_response) <- "double"
prc_response_hellinger <- vegan::decostand(
  prc_response, method = "hellinger"
)

prc_model <- vegan::prc(
  prc_response_hellinger,
  treatment = wave2b_m010_prc$treatment,
  time = wave2b_m010_prc$time
)

prc_permutations <- permute::how(
  nperm = params$nperm,
  within = permute::Within(type = "none"),
  plots = permute::Plots(
    strata = factor(wave2b_m010_prc$population_sample_id),
    type = "free"
  ),
  blocks = wave2b_m010_prc$seed_block_id
)

set.seed(params$random_seed)
prc_test <- anova(
  prc_model,
  permutations = prc_permutations,
  first = TRUE
)
prc_test
```

`first = TRUE` tests the first PRC axis, the axis shown in the usual PRC plot.
The official [`anova.cca` documentation](https://vegandevs.github.io/vegan/reference/anova.cca.html)
describes global, axis, term, and marginal permutation tests.

## 8.3 Plot and interpret the PRC

```{r plot-prc, eval=params$run_models, fig.cap="Principal response curves for host-return treatments relative to the matched no-return trajectory at m = 0.1. Only the ancestral lineages with the largest absolute weights are labelled."}
prc_lineage_scores <- vegan::scores(
  prc_model, display = "species", choices = 1
)
top_lineages <- order(abs(prc_lineage_scores[, 1]), decreasing = TRUE)[1:12]

plot(
  prc_model,
  select = top_lineages,
  legpos = "topright",
  xlab = "Host-population passage",
  ylab = "Treatment deviation from the no-return trajectory"
)
```

Read this figure in two parts:

- a treatment curve at zero resembles the reference community after removing
  changes shared by every treatment at that passage;
- a curve farther from zero shows a larger multivariate departure;
- ancestral lineages with large positive weights tend to follow the direction
  of a positive treatment curve, and lineages with large negative weights tend
  to move oppositely;
- the overall sign of a constrained axis is arbitrary, so reversing every sign
  would not change the biological result.

Do not interpret a single curve peak as a separate significant event unless a
time-specific contrast was prespecified. The formal PRC test evaluates the
multivariate trajectory contrast as a whole.

# 9. Save a reproducible result bundle

Save models, permutation-test tables, parameter settings, and session details.
This avoids relying on figures alone and makes the exact analysis auditable.

```{r save-results, eval=params$run_models}
result_bundle <- list(
  parameters = params,
  endpoint_fits = fits,
  wave1_dbrda_tests = wave1_db_tests,
  wave1_hellinger_tests = wave1_hellinger_tests,
  prc_comparison = unique(wave2b_m010_prc[
    , c("cell", "treatment", "H", "alpha_target", "m", "u")
  ]),
  prc_model = prc_model,
  prc_test = prc_test,
  session_info = sessionInfo()
)

saveRDS(
  result_bundle,
  file.path(analysis_output, "phase1-stage3-community-analysis.rds")
)
capture.output(
  sessionInfo(),
  file = file.path(analysis_output, "R-session-info.txt")
)
```

# 10. Reporting checklist

For every reported analysis, state:

1. the analysis set and exact cells included;
2. whether the response was total-variation distance or Hellinger-transformed
   ancestral-lineage frequency;
3. the model formula and any distance correction;
4. that seed block was conditioned and permutations were restricted within
   seed block;
5. the number of population replicates and permutations;
6. the global test, relevant predeclared term/interaction tests, effect size,
   and adjusted explained variation;
7. for PRC, the reference treatment and confirmation that whole trajectories,
   rather than individual passages, were permuted;
8. the software versions and random seed.

If several PRCs or planned contrasts are formally tested, specify the family of
tests in advance and adjust for multiplicity (for example, Holm correction).
Exploratory curves can still be shown, but they should be labelled exploratory.
